\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-22 paper.qmd
Add selected three parameter settings for PBMC example

@@

@@

>>>>

- The most important parameter is n_neighbors - the number of approximate nearest neighbors used to construct the initial high-dimensional graph. It effectively controls how UMAP balances local versus global structure - low values will push UMAP to focus more on local structure by constraining the number of neighboring points considered when analyzing the data in high dimensions, while high values will push UMAP towards representing the big-picture structure while losing fine detail.

- The second parameter we’ll investigate is min_dist, or the minimum distance between points in low-dimensional space. This parameter controls how tightly UMAP clumps points together, with low values leading to more tightly packed embeddings. Larger values of min_dist will make UMAP pack points together more loosely, focusing instead on the preservation of the broad topological structure.

@@

<!--UMAP parm 2-->

#tSNE_pbmc <- read_rds("data/pbmc/pbmc_tsne_30.rds")
UMAP_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_umap_5_min_dist_0.1.rds")
# PHATE_pbmc <- read_rds("data/pbmc/pbmc_phate.rds")
# TriMAP_pbmc <- read_rds("data/pbmc/pbmc_trimap.rds")
# PaCMAP_pbmc <- read_rds("data/pbmc/pbmc_pacmap.rds")

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)
benchmark_pbmc <- 1.055271

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.1) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/SO3thJ8PMi8>)"
#| label: fig-nldervisPBMCUMAP2
 #| out-width: 100%

<!--UMAP parm 3-->

#tSNE_pbmc <- read_rds("data/pbmc/pbmc_tsne_30.rds")
UMAP_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_umap_30_min_dist_0.01.rds")
# PHATE_pbmc <- read_rds("data/pbmc/pbmc_phate.rds")
# TriMAP_pbmc <- read_rds("data/pbmc/pbmc_trimap.rds")
# PaCMAP_pbmc <- read_rds("data/pbmc/pbmc_pacmap.rds")

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)
benchmark_pbmc <- 1.057549

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.01) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/hP8ie_oMXw0>)"
#| label: fig-nldervisPBMCUMAP3
 #| out-width: 100%

@@

- The parameter is, in a sense, a guess about the number of close neighbors each point has.

### 2024-01-22 paper.tex
Add selected three parameter settings for PBMC example

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\end{itemize}

@@

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
\end{itemize}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\item
  The most important parameter is n\_neighbors - the number of
  approximate nearest neighbors used to construct the initial
  high-dimensional graph. It effectively controls how UMAP balances
  local versus global structure - low values will push UMAP to focus
  more on local structure by constraining the number of neighboring
  points considered when analyzing the data in high dimensions, while
  high values will push UMAP towards representing the big-picture
  structure while losing fine detail.
\item
  The second parameter we'll investigate is min\_dist, or the minimum
  distance between points in low-dimensional space. This parameter
  controls how tightly UMAP clumps points together, with low values
  leading to more tightly packed embeddings. Larger values of min\_dist
  will make UMAP pack points together more loosely, focusing instead on
  the preservation of the broad topological structure.
\end{itemize}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP2-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP2}(a) 2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.1) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/SO3thJ8PMi8})}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP3-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP3}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.01) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/hP8ie_oMXw0})}

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\end{itemize}
